Disease and climate effects on individuals drive post-reintroduction

### 2018-08-30 article.Rmd
working on revisions based on reviews

---
title: Disease and climate effects on individuals jointly drive post-reintroduction
  population dynamics of an endangered amphibian
bibliography: library.bib
output:
    pdf_document:
      keep_tex: true
      includes:
          in_header: header.sty
fontsize: 12pt
csl: ecology.csl
---

@@

\begin{abstract}
The emergence of novel pathogens often has dramatic negative effects on previously unexposed host populations. 
Subsequent disease can drive populations and even species to extinction. 
After establishment in populations, pathogens can continue to affect host dynamics, influencing the success or failure of species recovery efforts. 
However, quantifying the effect of pathogens on host populations in the wild is challenging because individual hosts and their pathogens are difficult to observe. 
Here we use long-term mark-recapture data to describe the dynamics of reintroduced populations of an endangered amphibian (\textit{Rana sierrae}) and evaluate the success of these recovery efforts in the presence of a recently-emerged pathogen, the amphibian chytrid fungus \textit{Batrachochytrium dendrobatidis}. 
We find that high \textit{B. dendrobatidis} infection intensities are associated with increases in detectability, reductions in survival, and more infected adults. 
We also find evidence for intensity-dependent survival, with heavily infected individuals suffering higher mortality. 
These results highlight the need in disease ecology for probabilistic approaches that account for uncertainty in infection intensity using imperfect observational data. 
Such approaches can advance the understanding of disease impacts on host population dynamics, and in the current study will improve the effectiveness of species conservation actions.

@@

Assuming that survival depends only on whether an individual is infected, and not considering infection intensity (i.e., load), could be problematic given that Bd-caused frog mortality is often a function of load [@briggs2010enzootic]. 
However, individual-level infection data are often not available. 
For example, Bd load may not be observed if an individual is not encountered or captured on a survey, or if they are captured but no infection data are collected. 
This presents a major challenge: infection intensity is hard to measure, but may be critical for understanding population dynamics. 
Previous efforts to understand disease impacts in wild populations as a function of load at the individual level have randomly imputed missing infection load data with the observed distribution of load values [@spitzen2017fragile]. 
However, random imputation could bias inference if load data are not missing at random, for example, if load-dependent disease affects detection.

In this paper, we use a Bayesian multi-state model with data from decade-long mark-recapture studies in two reintroduced frog populations to understand the influence of disease and climate on introduction outcomes. 
We find that disease dynamics affect introduction success, with high infection intensities leading to more infected adults, a reduction in adult survival, and adults that are easier to detect. Recruitment and uninfected adult survival are also important for population persistence. Differential detectability of adults related to infection intensity points to broader questions about understanding disease impacts in wild populations, and highlights the importance of understanding hidden infection dynamics at the individual level. 
These insights will inform ongoing and future efforts aimed at restoring the  endangered mountain yellow-legged frog, and provide a means to quantitatively assess why some introduction efforts fail and others succeed.

@@

The initial introduction of frogs to Alpine and Subalpine occurred during summer 2006 and 2008, respectively, and additional introductions to supplement both populations were conducted in subsequent years. 
In the first introduction to Alpine, frogs were transported on foot; in all subsequent introductions to both Alpine and Subalpine, frogs were transported via helicopter (Table \ref{tab:table1}).

@@

We developed a hierarchical Bayesian hidden Markov model to understand how environmental factors and partly observed time-varying individual traits (Bd loads) jointly drive population dynamics. 
This model also needed to account for the partly deterministic recruitment process that arises with introductions. 
The result is an extended open population Jolly-Seber mark-recapture model with known additions to the population (introduced adults), continuous uncertain disease states of individuals which may or may not be captured, and sampling that is unevenly distributed in time.

@@

total_swabs <- alpine_n + subalpine_n
prop_miss <- (alpine_n_missing + subalpine_n_missing) / total_swabs
pct_miss <- 100 * prop_miss
```

In a small number of cases, adults were captured, weighed, and measured, but no swab data were associated with the capture (e.g., if a swab was lost). 
In these cases (`r round(pct_miss, 1)`\% of `r total_swabs` swabs), capture data indicate that the adult was alive during the survey, but its infection status is unknown. 
As such, there are three possible observations for captured frogs: detected/infected, detected/uninfected, and detected/unknown infection status. 
A fourth possible observation class corresponds to non-detection: an individual was not observed during a survey.

@@

The fact that individuals are imperfectly detected, and that occasionally, individuals are detected but no disease information is recorded presents a challenge for including individual-level Bd loads in the model. 
Within the infected population, Bd loads are partially observed when individuals are captured and swabs are collected. 
We used a normal distribution to represent (potential) log Bd loads:

@@

@@

@@

Despite the two study populations being 12 km apart, mean Bd loads among infected adults varied synchronously between populations. 
After initial introduction in 2006 or 2008, mean loads at both sites were typically between 1,000 and 10,000 copies, with a reduction in mean Bd load during the year following the first introduction (Figure \ref{mean-bd-loads}). 
Thereafter, loads were relatively low and stable through 2012. 
At both sites mean loads were uncharacteristically low in 2013 but increased in subsequent years until reaching a peak in 2016. 
Mean Bd loads at Subalpine tended to be higher than at Alpine, particularly in the period from 2015 - 2017. 
The estimated correlation over time between mean log Bd loads at the two study sites was `r median(unlist(bd_load_correlations)) %>% round(n_digit)` (`r quantile(unlist(bd_load_correlations), .05) %>% round(n_digit)`, `r quantile(unlist(bd_load_correlations), .95) %>% round(n_digit)`). 
This correlation was not due to a shared response to winter severity because winter severity did not influence Bd load at either site (at Alpine $\beta^{(\mu)}$: `r beta_z_df %>% filter(site == "Alpine") %>% select(med) %>% df2num` (`r beta_z_df %>% filter(site == "Alpine") %>% select(lo) %>% df2num`, `r beta_z_df %>% filter(site == "Alpine") %>% select(hi) %>% df2num`), at Subalpine $\beta^{(\mu)}$: `r beta_z_df %>% filter(site == "Subalpine") %>% select(med) %>% df2num` (`r beta_z_df %>% filter(site == "Subalpine") %>% select(lo) %>% df2num`, `r beta_z_df %>% filter(site == "Subalpine") %>% select(hi) %>% df2num`).

@@

At both sites, transitioning from the uninfected to the infected adult class (gaining infection) was more likely than transitioning from the infected to uninfected class (losing infection) over most of the study period. 
Among-year variation in transition probabilities was similar at both sites, with 2013 having higher than average loss-of-infection probabilities (Figure \ref{transition-plot}A). 
This was most likely due to exceptionally low mean Bd loads in 2013: when Bd loads were low, adults were more likely to lose infection at both study sites, and less likely to gain infection in the Subalpine population (Figure \ref{transition-plot}B).

@@

The modeling approach developed here provides a quantitative means to assess how climate and disease affect demographic rates and population dynamics in hard-to-sample host populations. 
This builds on previous research that incorporated individual-level covariates in mark-recapture models [@pledger2003open; @royle2008modeling; @gimenez2010individual;  @ford2012incorporating], and that evaluated effects of Bd load at the individual level in wild populations [@spitzen2017fragile]. 
Our approach is unique in part because we treat unobserved infection status and intensity as parameters, rather than using imputation to backfill missing intensity values. 
Accounting for uncertainty in individual-level infection intensity in a Bayesian framework simplifies uncertainty propagation to key parameters including detection and survival probabilities. 
This may provide a better understanding of how individual traits relate to demographic processes and population dynamics, especially when effects of disease on detectability lead to biased capture histories.

@@

The two study lakes were relatively distant from each other and located in catchments with different environmental characteristics, but despite this Bd loads were temporally correlated across the two sites. 
The synchrony in expected load is worth further investigation, as it may indicate environmental forcing of disease dynamics, e.g., variation due to temperature [@cohen2017thermal; @phillott2013chytridiomycosis], but see [@knapp2011nowhere], or some form of connectivity among sites, although the latter possibility seems unlikely. 
Future studies might also seek to better disentangle landscape-wide factors driving synchrony and local factors that drive differences in mean load, particularly in the context of known mechanisms of disease-induced extinction such as non-density dependent transmission [@rachowicz2007quantifying; @orlofske2018experimental], nonamphibian reservoir hosts [@mcmahon2013chytrid], and small equilibrium densities in the presence of disease [@de2005mechanisms].

@@

The recruitment model developed here may be somewhat simplistic in its focus on adults. 
In reality, recruitment into the adult population is a function of dynamics in the subadult and larval populations.
In particular, in mountain yellow-legged frogs and many other frog species Bd infection imposes heavy mortality during metamorphosis, but disease effects on larval stages are relatively weak [@rachowicz2004transmission; @rachowicz2006emerging]. 
Therefore, to fully understand how factors such as winter severity and disease influence population dynamics by way of affecting larvae, subadults, and adults, future efforts might focus on incorporating elements of integral projection models [@wilber2016integral] and integrated population models [@schaub2011integrated] with hidden Markov models similar to those we developed in this study.
A joint model for larval, subadult, and adult population dynamics would be better able to account for time lags resulting from the 1-4 year larval development period of this species [@Vredenburg2005mountain]. 
For example, one severe winter could have long-lasting effects if it slows the development of larvae so that they develop in three rather than two years, but this effect might not be detectable as a signal for adult recruitment until four years after the severe winter, depending on the growth rate of subadult frogs.

Our results provide important insights into causes of population establishment or likely failure for the endangered \textit{R. sierrae}. 
The two study populations showed very different patterns of adult survival and recruitment, and in particular, within a population the estimated survival of introduced cohorts was remarkably consistent across all introductions. 
Although the generality of results obtained from these two populations need to be assessed using data from mark-recapture efforts at other reintroduced populations, the results suggest that the estimated survival of reintroduced frogs could provide an early indication of the site-specific probability of introduction success. 
In addition, in the future, when survival estimates of reintroduced frogs are available from a larger number of populations, these estimates could eventually allow the identification of site characteristics associated with likely introduction success or failure. 
This predictive ability would greatly increase the effectiveness of mountain yellow-legged frog recovery efforts.

>>>>

---
title: Disease and climate effects on individuals drive post-reintroduction
  population dynamics of an endangered amphibian
bibliography: library.bib
output:
    pdf_document:
      keep_tex: true
      includes:
          in_header: header.sty
fontsize: 12pt
csl: ecology.csl
---

@@

\begin{abstract}
The emergence of novel pathogens often has dramatic negative effects on previously unexposed host populations. 
Subsequent disease can drive populations and even species to extinction. 
After establishment in populations, pathogens can continue to affect host dynamics, influencing the success or failure of species recovery efforts. 
However, quantifying the effect of pathogens on host populations in the wild is challenging because individual hosts and their pathogens are difficult to observe. 
Here we use long-term mark-recapture data to describe the dynamics of reintroduced populations of an endangered amphibian (\textit{Rana sierrae}) and evaluate the success of these recovery efforts in the presence of a recently-emerged pathogen, the amphibian chytrid fungus \textit{Batrachochytrium dendrobatidis}. 
We find that high \textit{B. dendrobatidis} infection intensities are associated with increases in frog detectability and reductions in survival. 
When average infection intensities are high, adults are more likely to gain infections and less likely to lose infections.
We also find evidence for intensity-dependent survival, with heavily infected individuals suffering higher mortality. 
These results highlight the need in disease ecology for probabilistic approaches that account for uncertainty in infection intensity using imperfect observational data. 
Such approaches can advance the understanding of disease impacts on host population dynamics, and in the current study will improve the effectiveness of species conservation actions.

@@

Assuming that survival depends only on whether an individual is infected, and not considering infection intensity (i.e., load), could be problematic given that Bd-caused frog mortality is often a function of load [@briggs2010enzootic]. 
However, individual-level infection data are often not available. 
For example, Bd load may not be observed if an individual is not encountered or captured on a survey, or if they are captured but no infection data are collected. 
This presents a major challenge: infection intensity is hard to measure, but may be critical for understanding population dynamics. 
Previous efforts to understand disease impacts in wild populations as a function of load at the individual level have randomly imputed missing infection load data with the observed distribution of load values [@spitzen2017fragile]. 
However, random imputation could bias inference if load data are not missing at random, for example, if heavily infected individuals are easier or more difficult to detect than individuals with lower loads.

In this paper, we seek to understand whether introduction outcomes can explained by post-introduction disease dynamics or climatic conditions. 
We expected that failed introductions might be attributed to some combination of disease dynamics that inhibit population growth and climatic conditions that do not favor adult survival or recruitment.
We use a Bayesian multi-state model with data from decade-long mark-recapture studies in two reintroduced frog populations.
We find that Bd affects introduction success via impacts on adult survival.
Winter severity also appears to regulate adult survival and recruitment.
In addition, we find that infected adults are easier to detect, which raises broader questions about understanding disease impacts in wild populations, and highlights the importance of accounting for hidden individual-level infection dynamics. 
These insights will inform ongoing and future efforts aimed at restoring the  endangered mountain yellow-legged frog, and provide a means to quantitatively assess why some introduction efforts fail and others succeed.

@@

The initial introduction of frogs to Alpine and Subalpine occurred during summer 2006 and 2008, respectively, and additional introductions to supplement both populations were conducted in subsequent years. 
In the first introduction to Alpine, frogs were transported on foot; in all subsequent introductions to both Alpine and Subalpine, frogs were transported via helicopter. 
The number of individuals introduced to each study site on each introduction date is provided in Table \ref{tab:table1}.

@@

We developed a hierarchical Bayesian hidden Markov model to understand how environmental factors and partly observed time-varying individual traits (Bd loads) jointly drive population dynamics. 
This model also needed to account for the partly deterministic recruitment process that arises with introductions. 
The result is an extended open population Jolly-Seber mark-recapture model with known additions to the population (introduced adults), continuous uncertain infection states of individuals which may or may not be captured, and sampling that is unevenly distributed in time.

@@

total_swabs <- alpine_n + subalpine_n
n_missing <- alpine_n_missing + subalpine_n_missing
prop_miss <- (n_missing) / total_swabs
pct_miss <- 100 * prop_miss
```

In a small number of cases, adults were captured, weighed, and measured, but no swab data were associated with the capture (e.g., if a swab was lost). 
In these cases (`r n_missing` of `r total_swabs` swabs), capture data indicate that the adult was alive during the survey, but its infection status is unknown. 
As such, there are three possible observations for captured frogs: detected/infected, detected/uninfected, and detected/unknown infection status. 
A fourth possible observation class corresponds to non-detection: an individual was not observed during a survey.

@@

The fact that individuals are imperfectly detected, and that occasionally, individuals are detected but no infection data are collected presents a challenge for including individual-level Bd loads in the model. 
Within the infected population, Bd loads are partially observed when individuals are captured and swabs are collected. 
We used a normal distribution to represent (potential) log Bd loads:

@@

## Detection probabilities

@@

## Infection dynamics

@@

Despite the two study populations being 12 km apart, mean Bd loads among infected adults varied synchronously between populations. 
After initial introduction in 2006 or 2008, mean loads at both sites were typically between 1,000 and 10,000 copies, with a reduction in mean Bd load during the year following the first introduction (Figure \ref{mean-bd-loads}A). 
Thereafter, loads were relatively low and stable through 2012. 
At both sites mean loads were uncharacteristically low in 2013 but increased in subsequent years until reaching a peak in 2016. 
Mean Bd loads at Subalpine tended to be higher than at Alpine, particularly in the period from 2015 - 2017. 
The estimated correlation over time between mean log Bd loads at the two study sites was `r median(unlist(bd_load_correlations)) %>% round(n_digit)` (`r quantile(unlist(bd_load_correlations), .05) %>% round(n_digit)`, `r quantile(unlist(bd_load_correlations), .95) %>% round(n_digit)`). 
This correlation was not due to a shared response to winter severity because winter severity did not influence Bd load at either site (at Alpine $\beta^{(\mu)}$: `r beta_z_df %>% filter(site == "Alpine") %>% select(med) %>% df2num` (`r beta_z_df %>% filter(site == "Alpine") %>% select(lo) %>% df2num`, `r beta_z_df %>% filter(site == "Alpine") %>% select(hi) %>% df2num`), at Subalpine $\beta^{(\mu)}$: `r beta_z_df %>% filter(site == "Subalpine") %>% select(med) %>% df2num` (`r beta_z_df %>% filter(site == "Subalpine") %>% select(lo) %>% df2num`, `r beta_z_df %>% filter(site == "Subalpine") %>% select(hi) %>% df2num`).

```{r get-cor-prev}
cor_prev_z <- read_csv('out/cor_prev_z.csv')
```

At both sites, transitioning from the uninfected to the infected adult class (gaining infection) was more likely than transitioning from the infected to uninfected class (losing infection) over most of the study period. 
Among-year variation in transition probabilities was similar at both sites, with 2013 having higher than average loss-of-infection probabilities (Figure \ref{transition-plot}A). 
This was most likely due to exceptionally low mean Bd loads in 2013: when Bd loads were low, adults were more likely to lose infection at both study sites, and less likely to gain infection in the Subalpine population (Figure \ref{transition-plot}B). 
As a consequence, when mean Bd loads were high, prevalence also tended to be high (Figure \ref{mean-bd-loads}A). 
The correlation between prevalence and mean Bd loads among infected adults was estimated to be `r cor_prev_z %>% filter(site == "Alpine") %>% select(med) %>% df2num` (`r cor_prev_z %>% filter(site == "Alpine") %>% select(lo) %>% df2num`, `r cor_prev_z %>% filter(site == "Alpine") %>% select(hi) %>% df2num`) at Alpine and `r cor_prev_z %>% filter(site == "Subalpine") %>% select(med) %>% df2num` (`r cor_prev_z %>% filter(site == "Subalpine") %>% select(lo) %>% df2num`, `r cor_prev_z %>% filter(site == "Subalpine") %>% select(hi) %>% df2num`) at Subalpine.

## Population dynamics

@@

@@

The modeling approach developed here provides a quantitative means to assess how climate and disease affect demographic rates and population dynamics in hard-to-sample host populations. 
This builds on previous research that incorporated individual-level covariates in mark-recapture models [@pledger2003open; @royle2008modeling; @gimenez2010individual;  @ford2012incorporating], and that evaluated effects of Bd load at the individual level in wild populations [@spitzen2017fragile]. 
Our approach is unique in part because we treat unobserved infection status and intensity as parameters, rather than using imputation to backfill missing intensity values. 
Accounting for uncertainty in individual-level infection intensity in a Bayesian framework simplifies uncertainty propagation to key parameters including detection and survival probabilities. 
This may provide a better understanding of how individual traits relate to demographic processes and population dynamics, especially when capture histories are biased by individual infection loads and/or disease.

@@

The two study lakes were relatively distant from each other and located in catchments with different environmental characteristics, but despite this Bd loads were temporally correlated across the two sites. 
The synchrony in expected load is worth further investigation, as it may indicate environmental forcing of infection dynamics, e.g., variation due to temperature [@cohen2017thermal; @phillott2013chytridiomycosis], but see [@knapp2011nowhere], or some form of connectivity among sites, although the latter possibility seems unlikely. 
Future studies might also seek to better disentangle landscape-wide factors driving synchrony and local factors that drive differences in mean load, particularly in the context of known mechanisms of disease-induced extinction such as non-density dependent transmission [@rachowicz2007quantifying; @orlofske2018experimental], nonamphibian reservoir hosts [@mcmahon2013chytrid], and small equilibrium densities [@de2005mechanisms].

@@

The recruitment model developed here may be somewhat simplistic in its focus on adults. 
In reality, recruitment into the adult population is a function of dynamics in the subadult and larval populations.
In particular, in mountain yellow-legged frogs and many other frog species Bd infection imposes heavy mortality during metamorphosis [@rachowicz2004transmission; @rachowicz2006emerging]. 
Therefore, to fully understand how climate and disease influence population dynamics by way of affecting larvae, subadults, and adults, future efforts might focus on incorporating elements of integral projection models [@wilber2016integral] and integrated population models [@schaub2011integrated] with hidden Markov models similar to those we developed in this study.
A joint model for larval, subadult, and adult population dynamics would be better able to account for time lags resulting from the 1-4 year larval development period of this species [@Vredenburg2005mountain]. 
For example, one severe winter could have long-lasting effects if it slows the development of larvae so that they develop in three rather than two years, but this effect might not be detectable as a signal for adult recruitment until four years after the severe winter, depending on the growth rate of subadult frogs.

Our results provide important insights into causes of population establishment or likely failure for the endangered \textit{R. sierrae}. 
The two study populations showed very different patterns of adult survival and recruitment, and in particular, within a population the estimated survival of introduced cohorts was remarkably consistent across all introductions. 
Although the generality of results obtained from these two populations need to be assessed using data from mark-recapture efforts at other reintroduced populations, the results suggest that the estimated survival of reintroduced frogs could provide an early indication of the site-specific probability of introduction success. 
Such indications could inform efficient allocation of limited conservation resources toward efforts that are likely to lead to persistent populations. 
In the future, when survival estimates of reintroduced frogs are available from a larger number of populations, these estimates could eventually allow the identification of site characteristics associated with likely introduction success or failure. 
This predictive ability would greatly increase the effectiveness of mountain yellow-legged frog recovery efforts.

More broadly, this work points to introduction of amphibian populations decimated by Bd as a potentially viable conservation action. 
Some introductions in other systems have failed due to chytridiomycosis [@stockwell2008impact; @soorae2010global], but the successful introduction of \textit{R. sierrae} in Yosemite provides evidence that introductions can succeed even when Bd persists.
Given the ongoing efforts to captively breed and then reintroduce threatened amphibians [@brannelly2016chytrid; @klop2016we; @murphy2017history], it is imperative to quantify the effects of Bd on host populations using imperfect observational data.
It is our hope that the framework developed here can be leveraged to better understand factors determining introduction success, and to develop more effective conservation interventions.
